Gradual Soundness: Lessons from Static Python

### 2022-04-29 pj-7.1/response.txt

> * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * *
>
>
> Review #45C
> ===========================================================================
>
> [... clipped ...]
>
> Assessment and comments (for authors)
> -------------------------------------
> The language design mixing concrete and shallow types is interesting.
> It seems that the language has been accepted by engineers and used for
> a large software project.  Nevertheless I have a few reservations to
> accept the paper in its current form.
>
> The paper puts emphasis on "gradual soundness" at the beginning but,
> after reading the paper, I failed to see what it really was.  The
> phrase is never mentioned on page 3 and afterwards.  This is
> unfortunate because this is the phrase used even in the title!

We've updated the paper to mention "gradual soundness" in the following
sections: 2.3, 2.5, 6.3, and 8.

> It's not clear how faithfully the formal model presented in Section 4
> models the implementation.  In particular, C-Sub allows the type of an
> expression can be changed to Dyn (using the consistent subtyping) and
> Matr allows conversion from Dyn to any non-Dyn type.  So, the type
> system seems to give any expression (that references only declared
> variables, functions, classes) any type.  However, apparently, some
> expressions are rejected by the implementation.  I have no idea what's
> really going on.  (I tried to access the Redex model but it wasn't
> accessible.)

You're absolutely right, mixing C-Sub and Matr is bad because it lets any
term have any type. (Run-time checks would catch the bad casts, but still,
we want to detect errors statically.)

We've revised section 4.2.2 to explain how the typing rules differ from
the Redex model's algorithm and in particular how the model uses subtyping
and materialization.

TODO

### 2022-04-29 pj-7.1/paper.tex
pj: kc + r3 fixes
- use Dyn instead of Object in rules
- explain the "magic" with C-Sub and Matr
- add None <: Opt[T]
- move comment about booleans 0 and 1

%% TODO after submission, before camera-ready
%% - reword 'fields' to attribute/descriptor/property?
%% - 
%% - 
%% -

@@

    \begin{tabular}{l@{\quad}l}
    {Eval. Type $\spteval$} & {Run-time check} \\\hline
    $\sptoptional{\spteval_0}$ & {Accepts either the none value or values that match $\spteval_0$} \\
    $\sptrawpydict$ & {Accepts any Python dictionary} \\
    $\sptchkdict{\spteval_0}{\spteval_1}$ & {Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    $\sptdyn$ & {No check needed, accepts any value} \\
    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$\tnote{*}} \\
    $\sptnone$ & {Accepts the Python none value} \\
    $\sptint$ & {Accepts integers} \\
    $\sptbool$ & {Accepts booleans (the integers \code{0} and \code{1})} \\
    $\sptstr$ & {Accepts strings}
    \end{tabular}

@@

    \inferrule*[left=D-App]{
      \sptvardef{\spf_0}{\sptdyn} \in \sptenv
      \\\\
      \sptenv \wtexpr \spexpr_0 : \sptobject
    }{
      \sptenv \wtexpr \spapp{\spf_0}{\spexpr_0} : \sptdyn
    }

    %% TODO show dict read rules?
%    \inferrule*[left=D-Ref]{
%      \sptenv \wtexpr \spexpr_0 : \sptrawpydict
%      \\\\
%      \sptenv \wtexpr \spexpr_1 : \sptobject
%    }{
%      \sptenv \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
%    }
%
%    \inferrule*[left=CD-Ref]{
%      \sptenv \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
%      \\\\
%      \sptenv \wtexpr \spexpr_1 : \spteval_0
%    }{
%      \sptenv \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
%    }

    \inferrule*[lab=D-Set]{
      \sptenv \wtexpr \spexpr_0 : \sptrawpydict
      \\\\
      \sptenv \wtexpr \spexpr_1 : \sptobject
      \\\\
      \sptenv \wtexpr \spexpr_2 : \sptobject
    }{
      \sptenv \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
    }

@@

    \inferrule*[left=C-Sub]{
      \sptenv \wtexpr \spexpr_0 : \spteval_1
      \\\\
      \sptenv \wtsub \spteval_1 \spcompat \spteval_0
    }{
      \sptenv \wtexpr \spexpr_0 : \spteval_0
    }

    \inferrule*[left=Matr]{
      \sptenv \wtexpr \spexpr_0 : \spteval_1
      \\\\
      \spteval_1 \spmatr \spteval_0
      %% side condition: result type is not a primitive
    }{
      \sptenv \wtexpr \spexpr_0 : \spteval_0
    }

@@

@@

By contrast to our Redex model, the typing judgment in~\cref{f:eval-types} is
not syntax directed.
The rules do not describe a typechecking algorithm.
Instead these rules declare various conclusions and their preconditions to
illustrate how \SP{} reasons about types---dynamic or otherwise.

The first two rules are for function application.
A typed function may be applied to an argument that matches its domain type,
in which case it computes a value that matches its codomain type.
A dynamically-typed variable may be applied to any input (that matches the top
type $\sptobject$) to yield a dynamically-typed result.
The next rules illustrate writes to Python dictionaries and checked dictionaries.
A shallow Python dict may be updated with any kind of key and value.
By contrast, a concrete checked dict requires keys and values that match its type parameters.
Next we have two rules for classes:
\trule{F-Set} says that writes to a class field must match the declared field
type; and \trule{M-App} shows that typed methods impose similar constraints as
typed functions.

The final two rules, \trule{C-Sub} and \trule{Matr}, depend on auxiliary judgments
for consistent subtyping ($\spcompat$) and materialization ($\spmatr$).
Consistent subtyping relates type $\spteval_0$ to type $\spteval_1$ if they are
related by static subtyping ($\spsubteq$) or if
$\spteval_1$ is the dynamic type.
Materialization relates the dynamic type to any non-dynamic type.
Occurrences of the \trule{Matr} rule are downcasts that require a
run-time check.

The name \emph{materialization} and the rule \trule{Matr} are inspired by prior
work~\cite{clps-popl-2019}.
By itself, the judgment is merely an upside-down type precision relation~\cite{g-icfp-2013,sv-dls-2008}.
Combined with the typing rule, however, materialization is a concise way to find
where a well-typed program needs casts to ensure soundness.
For example, suppose that $\spf$ is a function from integers to the dynamic
type and that $\spx$ is a variable with the dynamic type.
An application $\spapp{\spf}{\spx}$ can satisfy the type $\sptstr$ using two materializations:

\medskip
\begin{mathpar}
  \inferrule*[left=Matr]{
    \inferrule*{
      \inferrule*[left=Matr]{
        \sptenv \wtexpr \spx : \sptdyn
        \\
        \sptdyn \spmatr \sptint
      }{
        \sptenv \wtexpr \spx : \sptint
      }
    }{
      \sptenv \wtexpr \spapp{\spf}{\spx} : \sptdyn
    }
    \quad
    \sptdyn \spmatr \sptstr
  }{
    \sptenv \wtexpr \spapp{\spf}{\spx} : \sptstr
  }

  \mbox{Where $\sptenv = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
\end{mathpar}
\medskip
Consequently, this derivation calls for two casts at the \trule{Matr} applications.

@@

Furthermore, our casts-via-materialization rule assumes that the
type checker has access to the whole program.
In practice, \SP{} approximates our ideal casts in two steps
to support interactions with unanalyzed Python modules.
First, \SP{} inserts casts in an overapproximate way:
\begin{itemize}
  \item
    typed functions and methods begin by checking all their arguments;
  \item
    typed classes check all field writes;
  \item
    concrete-typed dictionaries check all key and value writes; and
  \item
    typed code checks the results of function calls, references, and other elimination forms
    whenever there is a materialization from the dynamic type.
\end{itemize}
Second, the type-directed optimizer skips casts wherever a well-typed context
interacts with a typed value.
\Cref{s:impl} explains the optimizer in more detail.

@@

Although sound gradual types are rare in industry, there are at
least three other languages that provide them.
Dart 2 is a nominally-typed language with a dynamic type that is similar
to a top (object) type but statically allows all method calls~\cite{dart-types}.
C\# has a dynamic type that delays type checking until runtime~\cite{bmt-ecoop-2010}.
\JSPP{} is typed JavaScript that allows untyped values
using a catch-all \code{external} type~\cite{jspp}.
None of these four implement gradual types as envisioned by researchers~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006};
instead, they offer a pragmatic solution to interoperability problems.
%% Programmers can manually cast external values to a more precise type.
%% The non-immediate casts make copies (how does that work for functions?!)
%% also, how does JS++ protect arrays sent to JS?

>>>>

%% TODO after submission, before camera-ready
%% - double-check section 4, can we be more convincing about soundness / what it is?
%% - reword 'fields' to attribute/descriptor/property?
%% -

%% TODO for camera-ready
%% -

@@

    \begin{tabular}{l@{\quad}l}
    {Eval. Type $\spteval$} & {Run-time check} \\\hline
    $\sptoptional{\spteval_0}$ & {Accepts either the none value or values that match $\spteval_0$} \\
    $\sptrawpydict$ & {Accepts any Python dictionary} \\
    $\sptchkdict{\spteval_0}{\spteval_1}$ & {Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    $\sptdyn$ & {No check needed, accepts any value} \\
    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$\tnote{*}} \\
    $\sptnone$ & {Accepts the Python none value} \\
    $\sptint$ & {Accepts integers and booleans} \\
    $\sptbool$ & {Accepts booleans} \\
    $\sptstr$ & {Accepts strings}
    \end{tabular}

@@

    \inferrule*[left=D-App]{
      \sptvardef{\spf_0}{\sptdyn} \in \sptenv
      \\\\
      \sptenv \wtexpr \spexpr_0 : \sptdyn
    }{
      \sptenv \wtexpr \spapp{\spf_0}{\spexpr_0} : \sptdyn
    }

    %% TODO show dict read rules?
%    \inferrule*[left=D-Ref]{
%      \sptenv \wtexpr \spexpr_0 : \sptrawpydict
%      \\\\
%      \sptenv \wtexpr \spexpr_1 : \sptdyn
%    }{
%      \sptenv \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
%    }
%
%    \inferrule*[left=CD-Ref]{
%      \sptenv \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
%      \\\\
%      \sptenv \wtexpr \spexpr_1 : \spteval_0
%    }{
%      \sptenv \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
%    }

    \inferrule*[lab=D-Set]{
      \sptenv \wtexpr \spexpr_0 : \sptrawpydict
      \\\\
      \sptenv \wtexpr \spexpr_1 : \sptdyn
      \\\\
      \sptenv \wtexpr \spexpr_2 : \sptdyn
    }{
      \sptenv \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
    }

@@

    \inferrule*[left=C-Sub*]{
      \sptenv \wtexpr \spexpr_0 : \spteval_1
      \\\\
      \sptenv \wtsub \spteval_1 \spcompat \spteval_0
    }{
      \sptenv \wtexpr \spexpr_0 : \spteval_0
    }

    \inferrule*[left=Matr*]{
      \sptenv \wtexpr \spexpr_0 : \spteval_1
      \\\\
      \spteval_1 \spmatr \spteval_0
      %% side condition: result type is not a primitive
    }{
      \sptenv \wtexpr \spexpr_0 : \spteval_0
    }

@@

    \inferrule*{
    }{
      \sptenv \wtsub \sptnone \spsubt \sptoptional{\spteval_0}
    }

@@

By contrast to our Redex model, the typing judgment outlined in~\cref{f:eval-types}
\emph{does not describe} a typechecking algorithm.
It is rather a declarative set of rules chosen to illustrate two points:
how \SP{} reasons about data structure types and how it enforces promises made
via the dynamic type.
The key steps that the Redex model takes to turn these rules into an algorithm
is to remove the two coercion rules
(both are marked with asterisks: \trule{C-Sub*} and \trule{Matr*})
and to allow restricted coercions at elimination forms.
For example, a function application may apply either consistent subtyping
or materialization to its argument.
It cannot apply both kinds of coecion because that can be abused to delay
any static error until run-time.

The first two rules are for function application.
A typed function may be applied to an argument that matches its domain type,
in which case it computes a value that matches its codomain type.
A dynamically-typed variable may be applied to any input
to yield a dynamically-typed result.
The following rules illustrate writes to Python dictionaries and checked dictionaries.
A shallow Python dict may be updated with any kind of key and value.
By contrast, a concrete checked dict requires keys and values that match its type parameters.
Next we have two rules for classes:
\trule{F-Set} says that writes to a class field must match the declared field
type; and \trule{M-App} shows that typed methods impose similar constraints as
typed functions.

The final two rules, \trule{C-Sub*} and \trule{Matr*}, depend on auxiliary judgments
for consistent subtyping ($\spcompat$) and materialization ($\spmatr$).
Consistent subtyping relates type $\spteval_0$ to type $\spteval_1$ if they are
related by static subtyping ($\spsubteq$) or if
$\spteval_1$ is the dynamic type.
Materialization relates the dynamic type to any non-dynamic type.
Occurrences of the \trule{Matr*} rule are downcasts that require a
run-time check.

The name \emph{materialization} and the rule \trule{Matr*} are inspired by prior
work~\cite{clps-popl-2019}.
By itself, the judgment is merely an upside-down type precision relation~\cite{g-icfp-2013,sv-dls-2008}.
Combined with the typing rule, however, materialization is a concise way to find
where a well-typed program needs casts to ensure soundness.
For example, suppose that $\spf$ is a function from integers to the dynamic
type and that $\spx$ is a variable with the dynamic type.
An application $\spapp{\spf}{\spx}$ can satisfy the type $\sptstr$ using two materializations:

\medskip
\begin{mathpar}
  \inferrule*[left=Matr*]{
    \inferrule*{
      \inferrule*[left=Matr*]{
        \sptenv \wtexpr \spx : \sptdyn
        \\
        \sptdyn \spmatr \sptint
      }{
        \sptenv \wtexpr \spx : \sptint
      }
    }{
      \sptenv \wtexpr \spapp{\spf}{\spx} : \sptdyn
    }
    \quad
    \sptdyn \spmatr \sptstr
  }{
    \sptenv \wtexpr \spapp{\spf}{\spx} : \sptstr
  }

  \mbox{Where $\sptenv = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
\end{mathpar}
\medskip
Consequently, this derivation calls for two casts at the \trule{Matr*} applications.

@@

Furthermore, our casts-via-materialization rule assumes that the
type checker has access to the whole program.
In practice, \SP{} approximates our ideal casts in two steps
to support interactions with unanalyzed Python modules.
First, \SP{} inserts casts in an overapproximate way:
\begin{itemize}
  \item
    typed functions and methods begin by checking all their arguments;
  \item
    typed classes check all field writes;
  \item
    concrete-typed dictionaries check all key and value writes; and
  \item
    typed code checks the results of function calls, references, and other elimination forms
    whenever there is a materialization from the dynamic type.
\end{itemize}
Second, the type-directed optimizer skips these casts wherever
a well-typed context interacts with a typed value.
\Cref{s:impl} explains the optimizer in more detail.

@@

Although sound gradual types are rare in industry, there are at
least three other languages that provide them.
Dart 2 is a nominally-typed language with a dynamic type that is similar
to a top type but statically allows all method calls~\cite{dart-types}.
C\# has a dynamic type that delays type checking until runtime~\cite{bmt-ecoop-2010}.
\JSPP{} is typed JavaScript that allows untyped values
using a catch-all \code{external} type~\cite{jspp}.
None of these four implement gradual types as envisioned by researchers~\cite{st-sfp-2006,tf-dls-2006,mf-toplas-2009,gktff-sfp-2006};
instead, they offer a pragmatic solution to interoperability problems.
%% Programmers can manually cast external values to a more precise type.
%% The non-immediate casts make copies (how does that work for functions?!)
%% also, how does JS++ protect arrays sent to JS?
